Pop-up cycleways provide a peak into post COVID-19 transport futures

### 2020-05-12 article.Rmd
Update articles

```{r gsub, fig.cap="Illustration of geographic subsetting based on distance to a central point (Leeds city city centre in this case) rather than based on location within somewhat arbitrarily shaped city boundaries. Radii of 5 km, 8 km and 10 km are shown for reference (note that some roads within 10 km of the center are outside the regional boundary)."}
# nrow(r_central)
# nrow(rj)
square = stplanr::geo_bb(shp = city_centre_buffer)
m1 = tm_shape(r_main_region) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(region) + tm_borders() +
  tm_shape(square) + tm_borders(lty = 3) +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() 
# m1
m2 = tm_shape(r_main) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(city_centre_buffer) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_small) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_large) + tm_borders(lty = 2) +
  tm_shape(city_key_buffer) + tm_borders() +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
tmap_arrange(m1, m2, nrow = 1)
```

>>>>

```{r gsub, fig.cap="Illustration of geographic subsetting based on distance to a central point (Leeds city centre in this case) rather than based on location within somewhat arbitrarily shaped city boundaries. Radii of 5 km, 8 km and 10 km are shown for reference (note that some roads within 10 km of the center are outside the regional boundary)."}
# nrow(r_central)
# nrow(rj)
square = stplanr::geo_bb(shp = city_centre_buffer)
m1 = tm_shape(r_main_region) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(region) + tm_borders() +
  tm_shape(square) + tm_borders(lty = 3) +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() 
# m1
m2 = tm_shape(r_main) +
  tm_lines("maxspeed", palette = "-Set1", lwd = "n_lanes_numeric", scale = 5,
           lwd.legend = c(1, 2, 3, 4, 5, 6), title.lwd = "Number of lanes") +
  tm_shape(city_centre_buffer) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_small) + tm_borders(lty = 2) +
  tm_shape(city_centre_buffer_large) + tm_borders(lty = 2) +
  tm_shape(city_key_buffer) + tm_borders() +
  tm_shape(h_leeds) + tm_dots(size = 0.5) +
  tm_scale_bar() +
  tm_layout(legend.show = FALSE)
tmap_arrange(m1, m2, nrow = 1)
```

### 2020-05-12 conversation.Rmd
Update articles

---
title: "Pop-up cycleways: a peak into post COVID-19 transport futures"
---

Around the world, as nations prepare to emerge from months of COVID-19 induced lockdown, life cannot simply return to previous norms. The pandemic has transformed all sectors of the economy, not least transport. Long distance travel has been dramatically altered, as [demand plummets](https://osf.io/preprints/socarxiv/v3g5d/) and [airline companies go bust](https://arxiv.org/abs/2004.08460). Yet short distance travel patterns have also changed greatly. Forced to stay close to home, many people have started walking and cycling more. From a health and well-being perspective, this may be a more important shift, especially in the long term as lockdown measures are gradually eased.

Active travel is already [recommended by the World Health Organisation](https://who.canto.global/pdfviewer/viewer/viewer.html?share=share%2Calbum%2CU6GDM&column=document&id=q3v02qdsh114763ffe3apo4s2p&suffix=pdf) as a safe mode of transport, and in the coming weeks and months it will play a vital role in enabling the reopening of many countries. As travel increases, we must ensure that safe, socially distanced walking and cycling remains possible. Maintaining social distancing on public transport could severely reduce the capacity of rail and bus networks, and there is a risk that this may lead to a rapid rise in road traffic. [Record bicycle sales](https://www.bbc.co.uk/news/business-52564351) suggest that many people have taken advantage of lockdown conditions to take up or return to cycling, but without careful management, increased traffic congestion could again deter many would-be cyclists from the streets.

Active travel has extensive [social and environmental benefits](https://www.bma.org.uk/what-we-do/population-health/preventing-ill-health/steps-to-increase-physical-activity-levels-in-the-uk). Meeting recommended levels of physical activity can help to prevent or manage 20 chronic health conditions, such as cancer, stroke and type 2 diabetes. It can manage or reduce the risk of mental health conditions such as depression and anxiety, and aid with maintaining a healthy weight. Evidence is beginning to emerge of potential links between air pollution and COVID-19 [infection](https://www.ncbi.nlm.nih.gov/pubmed/32315904) and [lethality](https://www.sciencedirect.com/science/article/pii/S0269749120320601), and many health conditions are [associated with poor air quality](https://www.ncbi.nlm.nih.gov/pubmed/30419237). Indeed, across Europe it is estimated that improvements in air quality in recent months have led to [11,000 fewer deaths](https://energyandcleanair.org/wp/wp-content/uploads/2020/04/CREA-Europe-COVID-impacts.pdf) from conditions such as these. This makes it all the more important to restrict any rise in road traffic and to encourage more beneficial means of travel.

Local and national governments worldwide are still working out how best to respond to these challenges. On Saturday 9th May, British Transport Secretary Grant Shapps announced a [£2 billion investment](https://www.gov.uk/government/news/2-billion-package-to-create-new-era-for-cycling-and-walking) in measures to promote walking and cycling. This follows a £10 million commitment from the Scottish Government and action announced in Wales. Hundreds of kilometres of temporary infrastructure is being put in place in dozens of cities worldwide, from [Milan](https://www.theguardian.com/world/2020/apr/21/milan-seeks-to-prevent-post-crisis-return-of-traffic-pollution) to [Martin Luther King Jr Drive in Philadelphia](https://www.theguardian.com/world/2020/apr/11/world-cities-turn-their-streets-over-to-walkers-and-cyclists ""), [Auckland](https://www.roadtraffic-technology.com/news/nz-govt-extra-wide-footpaths/) to [Mexico City](https://bicitekas.org/sites/default/files/public/manuales/propuesta_ciclovias_temporales_cdmxvfinal.pdf). Never has there been such a need for affordable, healthy and socially distanced transport that does not create the fumes that [emerging evidence] links to coronavirus susceptibility.

In terms of practical steps to encourage walking and cycling, a range of simple interventions can be undertaken, initially on a temporary basis. Roads can be closed to motor vehicles, or physical barriers can be used to enable [filtered permeability](https://cyclehighways.eu/design-and-build/infrastructure/filtered-permeability.html), in which access for residents is retained but modal filters enable routes for walking and cycling that are faster and more direct than the routes for motor vehicles. This is well established in [ the London Borough of Hackney ] where attractive barriers such as planters have been used to good effect. Another option involves keeping the roads open while reallocating excess roadspace to widen pavements or introduce pop-up cycleways.

These measures can all be undertaken on a temporary basis, but if action is prioritised in places where there is a *long-term need* for cycle improvements, there is a greater chance that people will demand they become permanent. This is what happened with the experimental cycleway in Torrington Place in London, where a 2 way road for motor vehicles has been converted into a thriving 2 way cycleway, enabling social distancing and zero emissions transport for keyworkers in central London. In Paris, measures aim to contribute towards Mayor Anne Hidalgo's long-term [Plan Velo](https://www.curbed.com/2020/1/15/21065343/bike-paris-cycling-anne-hidalgo).

With the rush to act, there is a great need for evidence of *where* new interventions should be prioritised. As with the medical science, research is needed now. Methods developed to identify locations of high walking and cycling potential can help ensure that the ‘pop-up’ infrastructure that goes in is effective, safe, and placed where it is most needed. Agile research can respond to rapidly changing conditions, as government priorities shift and different sectors of the economy reopen. Today the greatest need may be for medical workers to safely access [hospitals], tomorrow it could be access to schools or retail centres.

Building on the previous projects such as the [ Propensity to Cycle Tool ] and the [ Cycling Infrastructure Prioritisation Toolkit ], researchers from the University of Leeds have investigated where roadspace could be reallocated for pop-up cycleways. They have mapped out the top locations for intervention in ten major cities across England, with the intention of rolling out these methods across the country in the near future.

The methods have focused on road capacity and cycling potential, identifying roads where a ‘spare lane’ could be re-purposed for cycling. Cycle potential is assessed based on census data for travel to work and cycling uptake scenarios under the UK Government’s national target of doubling cycling levels by 2025. The length of road segments is also used to filter the top schemes.

The use of open data and publicly accessible models is a key aspect of this research. Given the global nature of the challenges we face, open and collaborative research is vital so that methods can be applied by as wide a user base as possible. The potential for international application can be seen for example in [ WHO report ] modelling cycling uptake in cities in low income countries. Within the UK, [ 50+ Local Authorities ] are now using the PCT to inform their strategic cycle network plans.

Maps (see below) show the recommendations for where to introduce pop-up cycleways. With the situation rapidly developing, some of the roads shown here, such as [Princess Street in Manchester](https://www.manchestereveningnews.co.uk/news/greater-manchester-news/part-deansgate-close-cars-city-18178175), have already been identified by Local Authorities as targets for intervention.

>>>>

---
title: "Pop-up cycleways provide a peak into post COVID-19 transport futures"
---

As nations prepare to emerge from months of COVID-19 induced lockdowns, it is becoming clear life cannot simply return to normal. The pandemic has transformed all sectors of the economy, not least transport.

In the coming weeks and months, walking and cycling will play a vital role in enabling the reopening of many countries, as continued social distancing measures will mean that public transport options are likely to run at a fraction of their intended capacity. There is a window of opportunity here that could be missed if roads become busy and dangerous again before new active transport habits become embedded.

Local and national governments are scrambling to act. In the UK, £2 billion investment in measures to promote walking and cycling has been announced. £250 million of this is for "emergency interventions to make cycling and walking safer" as [Transport Secretary Grant Shapps put it](https://www.gov.uk/government/speeches/transport-secretarys-statement-on-coronavirus-covid-19-9-may-2020). New [guidance from the Department for Transport](https://www.gov.uk/government/publications/reallocating-road-space-in-response-to-covid-19-statutory-guidance-for-local-authorities/traffic-management-act-2004-network-management-in-response-to-covid-19) advocates "measures to reallocate road space to people walking and cycling".

<tc-read-more id="137909"></tc-read-more>

Local authorities urgently need answers on how this funding should be spent, and where road space reallocation should be prioritised. New cyclinginfrastructure which is developed based on rigorous analysis of the best available data, in combination with local knowledge and community engagement, are more likely to be effective than plans based on intuition alone. Recognising the importance of a strong and actionable evidence-base, and the window of opportunity, we used our experience developing tools that are already being used by local authorities, notably the multi-institution PCT and the CyIPT, to develop methods and outputs taylored for the new transport planning landscape.

<image id="334100" align="centre" caption="Screenshot from the website salfordliveablestreets.commonplace.is to support local responses to the COVID-19 pandemic, including the prioritisation of pop-up active transport infrastructure." />

In a pre-print [paper](https://osf.io/7wjb6) on methods to prioritise pop-up road space reallocation schemes, we have classified pop-up interventions into three categories. Firstly, in streets that are wide or have multiple lanes, roadspace can be reallocated to widen pavements or introduce pop-up cycleways, such as outside a hospital in [Leicester](https://www.leicestermercury.co.uk/news/leicester-news/light-lockdown-traffic-leicester-allows-4083625). Secondly, physical barriers can be used to make routes for walking and cycling [faster and more direct](https://cyclehighways.eu/design-and-build/infrastructure/filtered-permeability.html) than the routes for motor vehicles. This can be seen in Tower Hamlets, London, where simple barriers like planters are used to good effect. Thirdly, roads can be closed altogether to motor vehicles, as is happening in [Deansgate, Manchester](https://www.manchestereveningnews.co.uk/news/greater-manchester-news/part-deansgate-close-cars-city-18178175).

The first of these three options is the most simple to implement,  with relatively few implications for traffic flow. We have mapped out **link** the top locations for this reallocation of road space in ten major cities across England, with the intention of rolling out these methods across the country in the near future. Having identified roads where a spare lane could be re-purposed for cycling, we have assessed the cycle potential of these streets based on census data for travel to work and cycling uptake scenarios under the UK Government’s national target of doubling cycling levels by 2025. The length of road segments is also used to filter the top schemes.

The use of open data and publicly accessible models is a key aspect of this research. Given the global nature of the challenges we face, open and collaborative research is vital so that methods can be applied by as wide a user base as possible. The potential for international application can be seen for example in [research for the World Health Organisation](https://atfutures.github.io/upthat/articles/adaptation.html) modelling cycling uptake in cities in low income countries. Within the UK, [50+ Local Authorities](https://npct.github.io/pct-shiny/regions_www/www/static/03d_other_reports/2019-use-of-pct-report.pdf) are now using the PCT to inform their strategic cycle network plans.

Results from our [methods](https://osf.io/7wjb6) show that many cities have 'spare lanes' under the current levels of traffic in city centres and near key worker destinations such as hospitals. These results could support the prioritisation of vital resources dedicated to pop-up infrastructure. With the situation rapidly developing, some of the roads shown here, such as [Kirkstall Road](https://www.transportxtra.com/publications/local-transport-today/news/65382/leeds-to-pilot-enhanced-cycling-safety-u) (in the top 10 results our methods generated for Leeds), have already been identified by Local Authorities as targets for intervention. As the recovery continues, we hope our methods will support cities to take the opportunities that the crisis presents for active travel.

<image id="334113" align="centre" caption="Roads in Leeds with at least one spare lane that could be reallocate to provide space for active travel, and locations of hospitals" />

<tweet url="https://twitter.com/laura_laker/status/1257660564098162689" />

With the rush to act, there is a great need for evidence of where new interventions should be prioritised. As with the medical science, research is needed now. Methods developed to identify locations of high walking and cycling potential can help ensure that the ‘pop-up’ infrastructure that goes in is effective, safe, and placed where it is most needed. Agile research can respond to rapidly changing conditions, as government priorities shift and different sectors of the economy reopen. Measures can be undertaken on a temporary basis, but if action is prioritised in places where there is a long-term need for cycle improvements, there is a greater chance that they become permanent. In Paris, Covid-19 related measures aim to contribute towards Mayor Anne Hidalgo's long-term [Plan Velo](https://www.curbed.com/2020/1/15/21065343/bike-paris-cycling-anne-hidalgo).
